Follow-up Samples

### 2017-02-20 submission/manuscript_outline_20161024.Rmd
discussion text corrections

From our results there were large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.  There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].  Although there was a detectable change in the carcinoma group and this change was towards what would be expected for normal controls, the follow up samples may not be a completely normal bacterial microbiome.

@@

We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).  What we found was that not only was the positive probability of a lesion reduced after lesion removal [Figure 3 & 4] but also that the commonly associated CRC bacteria were not highly represented within this model.  Rather, OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.  With only a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  We followed this up by creating a second model to classify initial versus follow up samples.  We found that this model was able to accurately classify initial versus follow up samples [Figure 3 & 4].  Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S3].

For the majority of tests performed there were no differences in the bacterial microbiome based on whether chemotherapy or radiation was received [Table S3].  Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).  These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.  For time differences, specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d].

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive ewithin our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 3C & 3D].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].

Many of the common OTUs between the different models used had many OTUs that taxonomically classified to potential butyrate producers [Table S2].  Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them.  Both butyrate and polyphenols are thought to be protective against cancer in part by reducing inflammation [@louis_gut_2014].   These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes.  One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].  High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].  Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].  Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.

@@

Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after surgical removal of lesion.  Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.  Our data provides evidence that commensal bacteria may be important in the development of polyps and also potentially the transition from adenoma to carcinoma.

>>>>

From our results there were large observed differences in the bacterial microbiome between pre and post lesion removal samples based on whether the individual had an adenoma or carcinoma.  There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure S1].

@@

We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).  What we found was that not only was the positive probability of a lesion reduced after lesion removal [Figure 3 & 4] but also that the commonly associated CRC bacteria were not highly represented within this model.  Rather, OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.  With only a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  We followed this up by creating a second model to classify initial versus follow up samples.  We found that this model was able to accurately classify initial versus follow up samples [Figure 3 & 4].  Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S2].  Although there was a detectable change in the carcinoma group and this change was towards what would be expected for normal controls, the follow up samples may not be a completely normal bacterial microbiome.

For the majority of tests performed there were no differences in the bacterial microbiome based on whether chemotherapy or radiation was received [Table S3].  Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).  These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.  Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d].

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 3C & 3D].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].

Many of the common OTUs between the two models used had many OTUs that taxonomically classified to potential butyrate producers [Table S2].  Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them.  Both butyrate and polyphenols are thought to be protective against cancer in part by reducing inflammation [@louis_gut_2014].   These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes.  One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].  High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].  Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].  Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.

@@

Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after adenoma or carcinoma removal.  Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.  Our data provides evidence that commensal bacteria may be important in the development of polyps and also potentially the transition from adenoma to carcinoma.

### 2017-02-20 submission/manuscript_outline_20161024.tex
discussion text corrections

From our results there were large observed differences in the bacterial
microbiome between pre and post surgery samples based on whether the
individual had an adenoma or carcinoma. There were much larger
differences between initial and follow up samples based on the thetayc
distance metric and in fecal blood as measured by FIT for individuals
with carcinoma versus adenoma {[}Figure 1{]}. However, there were no
differences between initial and follow up samples for Shannon Diversity,
observed OTUs, or evenness regardless of whether the individual had an
adenoma or carcinoma {[}Table S1{]}. There was also no differences in
relative abundance of any specific OTU for lesion, adenoma only, or
carcinoma only {[}Figure s1{]}. Although there was a detectable change
in the carcinoma group and this change was towards what would be
expected for normal controls, the follow up samples may not be a
completely normal bacterial microbiome.

@@

We next created a model that incorporated FIT and the bacterial
microbiome to be able to classify lesions (adenoma or carcinoma). What
we found was that not only was the positive probability of a lesion
reduced after lesion removal {[}Figure 3 \& 4{]} but also that the
commonly associated CRC bacteria were not highly represented within this
model. Rather, OTUs that made up the most important variables
overwhelmingly belonged to commensal bacteria. With only a single OTU
from a previously associated cancer bacterium (\emph{Porphyromonas
asaccharolytica}). We followed this up by creating a second model to
classify initial versus follow up samples. We found that this model was
able to accurately classify initial versus follow up samples {[}Figure 3
\& 4{]}. Providing additional information on the importance of commensal
bacteria was that there were a total of 23 OTUs in common to both models
and the vast majority belonged to regular residents of our gut community
{[}Table S3{]}.

For the majority of tests performed there were no differences in the
bacterial microbiome based on whether chemotherapy or radiation was
received {[}Table S3{]}. Within our study there was a significant
difference for the time elapsed in the collection of the follow up
sample between adenoma and carcinoma (uncorrected P-value \textless{}
0.05), with time passed being less for adenoma (253 +/- 41.3 days) then
carcinoma (351 +/- 102 days). These results would indicate that the
findings described were specific to the surgical intervention and that
some of the differences observed between carcinoma and adenoma samples
could be due to differences in collection time between samples for the
two different groups. For time differences, specifically, it could
confound the observation that carcinomas changed more than adenomas
{[}Figure 1a \& 1d{]}.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive ewithin our models. There are a number of reasons why this
may have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after surgery {[}Figure 3C \&
3D{]}. It should be noted that our study does not argue against the
importance of these CRC associated bacteria in the pathogenesis of
disease but rather that they are not the main bacteria changing after
surgical removal of lesion. In fact, it is possible that these CRC
associated bacteria are important in the transition from adenoma to
carcinoma and would be one explanation as to why in our data we not only
see high initial relative abundances in carcinoma and not adenoma
individuals but also large decreases in relative abundance in some of
those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

Many of the common OTUs between the different models used had many OTUs
that taxonomically classified to potential butyrate producers {[}Table
S2{]}. Another batch of OTUs classified to bacteria that can either
degrade polyphenols or are inhibited by them. Both butyrate and
polyphenols are thought to be protective against cancer in part by
reducing inflammation {[}13{]}. These protective compounds are derived
from the breakdown of fiber, fruits, and vegetables by resident gut
microbes. One example of this potential
diet-microbiome-inflammation-polyp axis is that \emph{Bacteroides},
which was highly prevalent in our models, are known to be increased in
those with high non-meat based protein consumption {[}14{]}. High
protein consumption in general has been linked with an increased CRC
risk {[}15{]}. Conversely, \emph{Bacteroides} are inhibited by
polyphenols which are derived from fruits and vegetables {[}16{]}. Our
data fits with the hypothesis that the microbial metabolites from
breakdown products within our own diet could not only help to shape the
existing community but also have an effect on CRC risk and disease
progression.

>>>>

From our results there were large observed differences in the bacterial
microbiome between pre and post lesion removal samples based on whether
the individual had an adenoma or carcinoma. There were much larger
differences between initial and follow up samples based on the thetayc
distance metric and in fecal blood as measured by FIT for individuals
with carcinoma versus adenoma {[}Figure 1{]}. However, there were no
differences between initial and follow up samples for Shannon Diversity,
observed OTUs, or evenness regardless of whether the individual had an
adenoma or carcinoma {[}Table S1{]}. There was also no differences in
relative abundance of any specific OTU for lesion, adenoma only, or
carcinoma only {[}Figure S1{]}.

@@

We next created a model that incorporated FIT and the bacterial
microbiome to be able to classify lesions (adenoma or carcinoma). What
we found was that not only was the positive probability of a lesion
reduced after lesion removal {[}Figure 3 \& 4{]} but also that the
commonly associated CRC bacteria were not highly represented within this
model. Rather, OTUs that made up the most important variables
overwhelmingly belonged to commensal bacteria. With only a single OTU
from a previously associated cancer bacterium (\emph{Porphyromonas
asaccharolytica}). We followed this up by creating a second model to
classify initial versus follow up samples. We found that this model was
able to accurately classify initial versus follow up samples {[}Figure 3
\& 4{]}. Providing additional information on the importance of commensal
bacteria was that there were a total of 23 OTUs in common to both models
and the vast majority belonged to regular residents of our gut community
{[}Table S2{]}. Although there was a detectable change in the carcinoma
group and this change was towards what would be expected for normal
controls, the follow up samples may not be a completely normal bacterial
microbiome.

For the majority of tests performed there were no differences in the
bacterial microbiome based on whether chemotherapy or radiation was
received {[}Table S3{]}. Within our study there was a significant
difference for the time elapsed in the collection of the follow up
sample between adenoma and carcinoma (uncorrected P-value \textless{}
0.05), with time passed being less for adenoma (253 +/- 41.3 days) then
carcinoma (351 +/- 102 days). These results would indicate that the
findings described were specific to the surgical intervention and that
some of the differences observed between carcinoma and adenoma samples
could be due to differences in collection time between samples for the
two different groups. Specifically, it could confound the observation
that carcinomas changed more than adenomas {[}Figure 1a \& 1d{]}.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after surgery {[}Figure 3C \&
3D{]}. It should be noted that our study does not argue against the
importance of these CRC associated bacteria in the pathogenesis of
disease but rather that they are not the main bacteria changing after
surgical removal of lesion. In fact, it is possible that these CRC
associated bacteria are important in the transition from adenoma to
carcinoma and would be one explanation as to why in our data we not only
see high initial relative abundances in carcinoma and not adenoma
individuals but also large decreases in relative abundance in some of
those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

Many of the common OTUs between the two models used had many OTUs that
taxonomically classified to potential butyrate producers {[}Table S2{]}.
Another batch of OTUs classified to bacteria that can either degrade
polyphenols or are inhibited by them. Both butyrate and polyphenols are
thought to be protective against cancer in part by reducing inflammation
{[}13{]}. These protective compounds are derived from the breakdown of
fiber, fruits, and vegetables by resident gut microbes. One example of
this potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}14{]}. High protein consumption in general has been linked with an
increased CRC risk {[}15{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}16{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression.
